---
title: "Project 3 connect to database"
author: "Keith DeNivo"
date: "2024-03-13"
output: html_document
---
```{r packages}
library(RODBC)
library(RSQLite)
library(RMySQL)
library(DBI)
library(odbc)
library(dplyr)
library(dbplyr)
library(RPostgres)
library(sqldf)
library(tidyverse)
```
Tried a lot of packages to connect to the database. I believe the only one required is the RMySQL. 

``` {r connection_to_db}

vectorpass <- c('t','h','e','q','u','i','c','k','b','r','o','w','n','f','o','x','j','u','m','p','o','v','e','r','l','a','z','y','d','o','g','0','1','2','3','4','5','6','7','8','9',".")
#list of alphanumeric characters to make the password into a simple code.

pass = paste(vectorpass[8],vectorpass[3],vectorpass[6],vectorpass[1],vectorpass[2],vectorpass[42],vectorpass[29],vectorpass[3],vectorpass[13],vectorpass[6],vectorpass[22],vectorpass[11],vectorpass[36],vectorpass[41],sep="")
#constructing the password from the list

mysqlconnection <- dbConnect(RMySQL::MySQL(),

                             host="cunydata607sql.mysql.database.azure.com",

                             dbname = "keith.denivo49",

                             port=3306,

                             user="keith.denivo49",

                             password=pass)
#connecting to database.
src_dbi(mysqlconnection)
#lists all the tables in the database with tbls.

```
the password is just keith.denivo49.  i copied and pasted what i did for assignment 2 because it was working.

```{r dataframe }

soft_skills_df <- dbReadTable(mysqlconnection, "soft_skills")
job_names_df <- dbReadTable(mysqlconnection, "job_names_untidy")
technical_skills_df <- dbReadTable(mysqlconnection, "technical_skills")
#read tables from database into a dataframe.
View(soft_skills_df)
View(job_names_df)
View(technical_skills_df)

```
puts each table into a data frame. i put the view so they can be seen.

# Job Names Tidying
```{r}

# Extracting the rows and columns from the job names data frame and putting them into 4 separate data frames 
job_names_one<- job_names_df[1:21, 1:4]
job_names_one<- job_names_one[-1, ]
colnames(job_names_one)<- c("Primary Key", "Job Posting Link", "Job Title", "Company")
job_names_two<- job_names_df[1:21, 5:8]
job_names_two<- job_names_two[-1, ]
colnames(job_names_two)<- c("Primary Key", "Job Posting Link", "Job Title", "Company")
job_names_three<- job_names_df[1:21, 9:12]
job_names_three<- job_names_three[-1, ]
colnames(job_names_three)<- c("Primary Key", "Job Posting Link", "Job Title", "Company")
job_names_four<- job_names_df[1:21, 13:16]
job_names_four<- job_names_four[-1, ]
colnames(job_names_four)<- c("Primary Key", "Job Posting Link", "Job Title", "Company")

# Combining the four separate data frames
jc<- rbind(job_names_one, job_names_two, job_names_three, job_names_four)

jn_df_final <- jc %>%
mutate("Platform" = str_extract(jc$`Job Posting Link`, "monster|linkedin|builtin|ziprecruiter"))


```


# Soft Skills Tidying
```{r}
#cleaning up some of the column names
colnames(soft_skills_df)[1] <- "Primary Key" 
colnames(soft_skills_df)[2] <- "communication" 
colnames(soft_skills_df)[3] <- "collaboration" 
colnames(soft_skills_df)[13] <- "detail oriented" 
colnames(soft_skills_df)[16] <- "cross functional" 
colnames(soft_skills_df)[18] <- "time management" 
colnames(soft_skills_df)[20] <- "fast paced" 
colnames(soft_skills_df)[28] <- "driven" 
colnames(soft_skills_df)[30] <- "critical thinking" 
```

```{r}
# Function to create new column excluding the first column
create_new_column <- function(row) {
  # Exclude the first column
  row <- row[-1]
  # If all elements in the remaining row are 0, return empty string
  if (all(row == 0)) {
    return("")
  } else {
    # Extract column names where the value is 1
    columns_with_ones <- colnames(soft_skills_df)[-1][as.logical(row)]
    # Concatenate column names with a comma separator
    return(paste(columns_with_ones, collapse = ","))
  }
}

# Apply the function to each row of the dataframe
new_column <- apply(soft_skills_df, 1, create_new_column)

# Add the new column to the dataframe
soft_skills_df$NewColumn <- new_column

# Print the modified dataframe
View(soft_skills_df)

```

```{r}
#filtering the columns to create a new dataframe
softskills_final <- select(soft_skills_df, "Primary Key", "NewColumn" )
colnames(softskills_final)[2] <- "Soft Skills"
```


# Soft Skills analysis
<b><header>Identify the most prevalent soft skill among the sampled job postings: </header></b>
```{r}
# Split the Soft Skills column into individual soft skills
softskills_split <- unlist(strsplit(softskills_final$ "Soft Skills", ","))

# Remove leading and trailing whitespaces
softskills_split <- trimws(softskills_split)

# Count the occurrences of each soft skill
soft_skills_counts <- table(softskills_split)

# Find the soft skill with the highest count
most_valued_soft_skill <- names(which.max(soft_skills_counts))

# Print the most abundant soft skill
print(most_valued_soft_skill)
```

Communication is the most prevalent soft skill among the job postings we sampled.

<b><header>Identfy the top 10 soft skills:</header></b>
```{r}
# Get the top 10 soft skills
top_10_soft_skills <- head(sort(soft_skills_counts, decreasing = TRUE), 10)

# Create the bar plot 
barplot(top_10_soft_skills, 
        col = rainbow(length(top_10_soft_skills)),
        main = "Top 10 Soft Skills",
        xlab = "Soft Skills",
        ylab = "Frequency",
        las = 2, 
        cex.names = 0.8,
        xaxt = "n")  # Remove x-axis labels

# Add a legend 
legend("topright", 
       legend = names(top_10_soft_skills), 
       fill = rainbow(length(top_10_soft_skills)),
       cex = 0.5,  # Set the font size of the legend
       ncol = 2)   # Set the number of columns in the legend

```

# Technical Skills Analysis

<b><header>Identify the most prevalent technical skill among the sampled job postings:</header></b>
```{r}
# Split the Technical Skills column into individual tech skills
technical_skills_split <- unlist(strsplit(technical_skills_df$ "Technical.Skills", ","))

# Remove leading and trailing whitespaces
technical_skills_split <- trimws(technical_skills_split)

# Convert all technical skills to lowercase to account for both "python" and "Python"
technical_skills_lower <- tolower(technical_skills_split)

# Count the occurrences of each technical skill
technical_skills_counts <- table(technical_skills_lower)

# Find the technical skill with the highest count
top_tech_skill <- names(which.max(technical_skills_counts))

# Print the most prevalent technical skill
print(top_tech_skill)
```

Python is the most prevalent technical skill among the job postings we sampled. 
<b><header>Identify the top 10 technical skills:</header></b>
```{r}
top_ten_tech_skills <- head(sort(technical_skills_counts, decreasing = TRUE), 10)

# Create the bar plot
barplot(top_ten_tech_skills,
        col = heat.colors(length(top_ten_tech_skills)),
        main = "Top 10 Most Prevalent Technical Skills",
        xlab = "Technical Skills",
        ylab = "Frequency",
        las = 2, 
        cex.names = 0.8,
         xaxt = "n")

# Create a legend 
legend("topright", 
       legend = names(top_ten_tech_skills), 
       fill = heat.colors(length(top_ten_tech_skills)),
       title = "Technical Skills",
       cex = 0.5)  
```

# Combine all data frames for a merged data frame anyone can use
```{r}
#Rename to match the other "Primary Key" column names
names(technical_skills_df)[names(technical_skills_df) == "Primary.Key"] <- "Primary Key"

# Rename the "Technical.Skills" column for consistency
names(technical_skills_df)[names(technical_skills_df) == "Technical.Skills"] <- "Technical Skills"

# Check the modified column names
print(names(technical_skills_df))
```


```{r}
# Merge the data frames based on the "Primary Key" columns
merged_df <- jn_df_final %>%
              merge(softskills_final, by = "Primary Key") %>%
              merge(technical_skills_df, by = "Primary Key")

```

```{r}
#The primary key is not organized in numerical order due to the objects being characters. Convert them to numeric data types for effective use in the future.
merged_df$`Primary Key` <- as.numeric(merged_df$`Primary Key`)

#Check col names
colnames(merged_df)
```
